---
title: "Exploring intron signal in RNA-seq data using coverage estimation"
author: "Stuart Lee, Charity Law"
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
                      fig.align = "center",
                      comment = "#>", 
                      message = FALSE,
                      warning = FALSE,
                      fig.path = here::here("img/"),
                      cache.path = here::here("cache/")
                      )

# pre load libraries
suppressPackageStartupMessages(library(plyranges))
suppressPackageStartupMessages(library(superintronic))

# useful functions
hex <- function(data, mapping, ...) {
  ggplot2::ggplot(data = data, mapping = mapping) + 
    ggplot2::geom_hex(...) +
    # ggplot2::geom_point(data = filter(data, gene_id %in% gene_paper), colour = "red") +
    ggplot2::scale_fill_viridis_c(direction = -1)
}

save_plot <- function(.x, .y) {
  target <- filter(parts_sub, gene_id == !!.y$gene_id)
  
  fname <- paste0(target$gene_name, "_coverage_plot.pdf")
  fname <- here::here("img", fname)
  p <- view_coverage(cvg_over_features, target, facets = "CellLine")
  print(p)
  ggplot2::ggsave(fname, p, dpi = "retina")
} 


skip <- !file.exists(here::here("data", "cvg_over_features.rds"))

```

# Introduction

This report looks finding intron retention genes using simple summary 
statistics and visualisations using `superintronic`.

## Data

We will use data from 12 samples from the human cell line mixture RNA-seq
data. There are three biological replicates across two library preparation kits,
a polyA kit and a total RNA kit. The BAM files have been previously aligned
with subread using the hg20 reference genome. 

```{r prepare-design, echo= FALSE}
design <- readRDS(here::here("data", "design.rds"))
```


## `superintronic` analysis

A superintronic analysis consists of four steps:
1. Preparing the annotation and extract exonic/intronic parts
2. Computing coverage over regions of interest
3. Summarising coverage over regions of interest
4. Visualising the results

Steps 1 and 2. are performed in the `scripts/run_coverage.R`

### Preparing GFF/GTF files

```{r gff}
library(superintronic)
parts <- readRDS(here::here("data", "complete-annotation.rds"))
parts
```

These can be interrogated as desired, but for our purposes we will keep
genes that are protein coding, on main chromosomes but not mitochondrial 
genome, and do not overlap any other genes and have more than one exon.

```{r default-filter, echo = FALSE}
parts_sub <- readRDS(here::here("data", "filtered-annotation.rds"))
parts_sub
```



## Coverage over intronic/exonic regions

Now all the ingredients are in place to merge the coverage scores
to our prepared annotation with `join_parts()`. This again
returns a GRanges object, restricted to the intersection of the 
coverage ranges with intron/exon ranges. Additional columns are added
corresponding to the properties of the intron/exon.


```{r, load-features, echo = FALSE }
cvg_over_features <- readRDS(here::here("data", "parts-coverage.rds"))
```

Now we will compute summaries (rangenostics) for each cell line and kit to detect IR 
genes. This consists of three steps

1. Computing the average log-coverage over all exon/intron features:

```{r rangenostics-01}
# compute intron/exon features
cvg_over_features <- cvg_over_features %>% 
  mutate(log_score = log2(score + 0.5))  
  
cvg_by_features <- cvg_over_features %>% 
  group_by(Kit, CellLine, gene_id, feature_type)

E_vals <- cvg_by_features %>% 
  summarise(mn = Hmisc::wtd.mean(log_score, width),
            sd = sqrt(Hmisc::wtd.var(log_score, width)),
            raw_mn = Hmisc::wtd.mean(score, width),
            score = score,
            n_bases = width,
            start = min(start),
            end = max(end),
            seqnames = unlist(unique(seqnames)),
            strand = unlist(unique(feature_strand))) %>% 
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)

```


2. For exon features computing the mean and variance of average log-coverage 
   over all genes (lumpiness).
   
```{r rangenostics-02}
lumpy <- E_vals %>% 
  filter(sd > 0, feature_type == "exon") %>% 
  group_by(Kit, CellLine) %>% 
  summarise(
    E_mu = mean(mn),
    E_sd = mean(sd),
    E_raw = mean(raw_mn)
)

lumpy 

mcols(E_vals) <- cbind(mcols(E_vals),
                       lumpy[match(paste0(E_vals$Kit, E_vals$CellLine), 
                                   paste0(lumpy$Kit, lumpy$CellLine)), -c(1,2)])

E_vals <- E_vals %>% 
  mutate(
    n_bases_E_raw = sum(n_bases[score > E_raw]),
  ) 
```
   
   
3. Compute the average number of bases with a 
   log-coverage score above the E values defined above. We have also computed
   a smooth between over the intron/exon feature standard deviation against
   the average for visualisation purposes.

```{r rangenostics-03}
rango <- E_vals %>% 
  plyranges::select(-n_bases, -score, -seqnames, -start, -end, -width, -strand, 
         .drop_ranges = TRUE) %>% 
  dplyr::as_tibble() %>% 
  group_by(Kit, CellLine, feature_type) %>% 
  tidyr::nest() %>% 
  mutate(
    smooth = lapply(data, function(x) {
      mgcv::gam(sd ~ s(mn), data = x)
      }),
    augment = lapply(smooth, broom::augment),
  ) %>% 
  tidyr::unnest(data, augment) %>% 
  dplyr::select(-dplyr::ends_with("1"), -.sigma)
```

4. Plot the results.

We get very similar results to a `voom` plot in a standard limma analysis. We
have also overlaid on the E_mu and E_sigma values as potential cut-offs for
each feature. 

```{r voom-like-plot}
# rangenostics feature_average vs feature_sd coloured by n of bases above
# average raw exon coverage
library(ggplot2)

voom_plot <- function(.x, .y) {
  p <- ggplot(data = .x, aes(x = mn, y = sd)) +
    geom_point() +
    geom_line(aes(y = .fitted), colour = "blue") +
    geom_vline(aes(xintercept = E_mu), 
               data  = dplyr::distinct(.x, feature_type, E_mu)) +
    geom_hline(aes(yintercept = E_sd),
               data = dplyr::distinct(.x, feature_type, E_sd)) +
    facet_wrap(~feature_type) +
    labs(
         subtitle = paste(.y$Kit, .y$CellLine),
         x = "mean log-coverage",
         y = "sd log-covreage"
    )
  fname <- here::here("img", 
                      paste0(.y$Kit, "-", .y$CellLine, 
                             "voom-like-plot.png")
  )
  ggsave(fname, p)

}

rango %>% 
  group_by(Kit, CellLine) %>% 
  dplyr::group_walk(voom_plot)

```

We also plot pairs of rangenositcs for each feature in a scatter plot matrix
(using a hexbin scatter plot). Again we get similar results to the plots
produced by a DE analysis.

```{r rangenostics-pairs}
all_features <- tidyr::gather(rango, "key", "value", 
                              -Kit, -CellLine, -gene_id, -feature_type, 
                              -E_mu, -E_sd, -E_raw) %>% 
  mutate(var = paste0(feature_type, "_", key)) %>% 
  plyranges::select(-feature_type, -key) %>% 
  arrange(gene_id) %>% 
  tidyr::spread(var, value)

pairs_plot <- function(.x, .y) {
  sub <- plyranges::select(.x, 
                           exon_mn, 
                           intron_mn, 
                           intron_sd, 
                           bases_above = intron_n_bases_E_raw)
  fname <- here::here("img", 
                      paste0(.y$Kit, "-", .y$CellLine, 
                             "-pairs-plot.png")
  )
  
  p <- GGally::ggpairs(sub, columns = 1:4, lower = list(continuous = hex))
  ggsave(fname, p)
}

all_features %>% 
  group_by(Kit, CellLine) %>% 
  dplyr::group_walk(pairs_plot)
  
  
```

5. Decide on cut-off over each feature. It appears that choosing genes
with high intron feature standard deviations, and with number of bases above
E_mu in the 99th percentile discriminates intron retention features well.

```{r rangenositcs-hits}
hits <- all_features %>% 
  group_by(Kit, CellLine) %>% 
  filter(exon_mn > E_mu, 
         intron_sd > E_sd, 
         intron_n_bases_E_raw > quantile(intron_n_bases_E_raw, 0.99))  
 
dplyr::count(hits)
```

6. We can also look at overlaps between sets:

```{r}
sets <- hits %>% 
  dplyr::distinct(gene_id) %>%
  arrange(gene_id) %>% 
  mutate(value = 1, var = paste(CellLine, Kit)) %>% 
  ungroup() %>%  
  plyranges::select(gene_id, var, value) %>% 
  tidyr::spread(var, value) %>% 
  dplyr::mutate_at(2:5, .funs = ~ dplyr::if_else(is.na(.), 0L, 1L)) 

sets <- as.data.frame(sets)
library(UpSetR)
pl <- upset(sets)
ggsave(here::here("img", "olaps-sets.png"))
```

```{r all-features-out, include = FALSE}
readr::write_rds(all_features, here::here("data", "superintronic-features.rds"))
readr::write_rds(hits, here::here("data", "superintronic-hits.rds"))
```

6. And show the coverage plots for the `nrow(hits)` genes.

```{r superintronic-coverage-plots}
hits %>% 
  dplyr::group_walk(save_plot)
```


